---
title: "Multivariate Statistical Methods"
subtitle: "Assignment 3"
author: "Allan Gholmi, Emma Wallentinsson, Rasmus Holm"
date: "`r Sys.Date()`"
fontsize: 10pt
geometry: margin=1in
output:
    pdf_document:
        toc: false
        number_sections: false
        fig_caption: true
---

```{r global-options, echo = FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_chunk$set(fig.pos='H', fig.align='center')
```

# Question 2

```{r}
library(psych)

data <- read.table("../data/T1-9.dat")
names(data) <- c("country", "100m", "200m", "400m", "800m", "1500m", "3000m", "marathon")
numeric_data <- data[, -1]
countries <- as.character(data$country)

S <- cov(numeric_data)
R <- cor(numeric_data)
factors <- 2
```

Since the data is measured in different units it is more appropriate to use the correlation matrix.

## Analysis on Covariance Matrix

```{r, fig.height=10, fig.width=6}
S_principal <- principal(S, factors, rotate="varimax", covar=TRUE)
S_factanalysis <- factanal(numeric_data, factors=factors, covmat=S, rotation="varimax")

S_factoranalysis_loadings <- S_factanalysis$loadings[, 1:2]
S_principal_loadings <- S_principal$loadings[, 1:2]

old <- par(mfrow=c(2, 1))
plot(S_factoranalysis_loadings, type="n", main="ML Factor Analysis")
text(S_factoranalysis_loadings, labels=names(numeric_data), cex=.7)

plot(S_principal_loadings, type="n", main="PCA")
text(S_principal_loadings, labels=names(numeric_data), cex=.7)
par(old)
```

```{r}
print("PCA")
S_principal$Vaccounted

print("FA")
S_factanalysis$loadings
```

We can see that the first principal component explains about 87\% of the variance and the largest loading is associated with the marathon which is clear from the plot.

```{r, fig.height=10, fig.width=6}
factor_scores <- factanal(numeric_data, factors=factors,
                          rotation="varimax", scores="regression")$scores
principal_scores <- principal(numeric_data, factors, scores=TRUE, covar=TRUE)$scores

old <- par(mfrow=c(2, 1))
plot(factor_scores, type="n", main="ML Factor Analysis")
text(factor_scores, labels=countries)

plot(principal_scores, type="n", main="PCA")
text(principal_scores, labels=countries)
par(old)
```

## Analysis on Correlation Matrix

```{r, fig.height=10, fig.width=6}
R_principal <- principal(R, factors, rotate="varimax", covar=FALSE)
R_factanalysis <- factanal(numeric_data, factors=factors, covmat=R, rotation="varimax")

R_factoranalysis_loadings <- R_factanalysis$loadings[, 1:2]
R_principal_loadings <- R_principal$loadings[, 1:2]

old <- par(mfrow=c(2, 1))
plot(R_factoranalysis_loadings, type="n", main="ML Factor Analysis")
text(R_factoranalysis_loadings, labels=names(numeric_data), cex=.7)

plot(R_principal_loadings, type="n", main="PCA")
text(R_principal_loadings, labels=names(numeric_data), cex=.7)
par(old)
```

```{r}
print("PCA")
R_principal$Vaccounted

print("FA")
R_factanalysis$loadings
```

Now the first two principal components explains about the same amount of variance and in total almost 92\% so its a pretty good fit. Similar values are true for the factors.


```{r, fig.height=10, fig.width=6}
factor_scores <- factanal(numeric_data, factors=factors,
                          rotation="varimax", scores="regression")$scores
principal_scores <- principal(numeric_data, factors, scores=TRUE, covar=FALSE)$scores

old <- par(mfrow=c(2, 1))
plot(factor_scores, type="n", main="ML Factor Analysis")
text(factor_scores, labels=countries)

plot(principal_scores, type="n", main="PCA")
text(principal_scores, labels=countries)
par(old)
```

Setting rotation to varimax means that the algorithm rotates the loadings such as to maximize their variances. As a result of this rotation, each variable loads more heavily on a single factor making the factors easier to interpret.

\newpage

# Appendix

## Code

```{r, eval=FALSE}

```
